Use of asynchronous circuits and leakiness in modeling biological circuit of the Caulobacter Crescentus cell cycle

### 2017-02-22 asyncleakyn.tex
big change

\section{Correct boolean model}
We expect biological systems to behave similarly to an asynchronous circuit because in biological systems there are no 'global clocks' that regulate the timing.  Because there is high variability between cells, and molecular interactions are of a stochastic nature, production from each gene element can have varying delay.  This variability differs from cell to cell and from one cycle to the next. Thus we would also expect a type of speed independence property to ensure that most of the time, the cells work reliably regardless of the random delays and advance through the cell cycle to produce new cells.
From the bottom up, we extract properties from biological networks and show how it is analogous to the well developed CMOS technology.  We start at the physics level and demonstrate certain properties so that we can group gene elements together and represent them with a higher level boolean model.  We then show how this boolean model can be abstracted into a higher level forming building blocks that we can use in an asynchronous circuit. 
In past papers, gene regulatory networks has been modeled as boolean circuits with each gene/protein element represented by a boolean gate.  Each gene logic gate has inputs, usually its activators and repressors, to determine whether the protein is being actively produced or not.  These are usually modeled as AND or OR gates with its promoter binding compounds as its only inputs.  However, in our investigation using the ODE model of each gene, we found that degradation is a crucial part of the logic gates as well.  Protein degradation can be sped up through controlling the production of degradation enzymes.  Meanwhile for stable proteins, it is possible for the base degradation rate, the half life of the protein to be equivalent to the length of the cell cycle, the only reduction in protein concentration coming from the splitting of cells.  Thus there needs to be a separate way of representing the production of proteins from its degradation.  We take inspiration from Complementary Metal–Oxide–Semiconductor (CMOS) technology.  In CMOS there is a pull up network and a pull down network.  In the inverter example, the pull up network consists of a PMOS and the pull down network consists of a NMOS.  PMOS behaves, on first order approximation, as a wire when the input voltage is low and as an open circuit when the input voltage is high.  NMOS is the opposite, it behaves as an open when the input voltage is low and as a wire when the input voltage is high.  Thus when the input voltage is low, the output voltage is pulled up by the PMOS to a high voltage.  Meanwhile when the input voltage is high, the output voltage is pulled down to ground by the NMOS.  The transcriptional and translational regulation of these genes are similar to a pull up network while the accelerated degradation is similar to a pull down network.  If the background degradation is significantly slower, this effect can be thought of as the parasitic capacitance that slowly leak the charge of the output signal.  Overtime it degrades the signal however for short time scales it retains its voltage value and is a dynamic memory element.
Many biological functions can be modeled with the Hill function, including mRNA production.  The sigmoidal shape of the Hill function is reminiscent of Voltage Transfer Characteristics of CMOS inverters.  As such, much of the design properties of CMOS circuits can be carried over.  The importance of the sigmoidal shape of the input vs output transfer curve during steady state allows for restoring property of the signal and hence unsullied propagation in its next stages.   In an ODE model of the protein production and degradation with the Hill function, the steady state solutions are the Hill function divided by the degradation constant.  In general the equation is of the form: $\frac{dprotein}{dt}=g(inputs)-protein\cdot\lambda$  where the function $g()$ is a sigmoidal curve such as the Hill function and $\lambda$ is the degradation constant.  This sigmoidal curve is then an input (the proteins that initiate transcription and production of the protein) versus the output, the current concentration of the protein.  If we treat each protein as its own gate, there exists similar restorative properties of the CMOS gate.

@@

One basic property thresolds must satisfy is to allow proper propagation of the signal.  The output of a gate should read as the same binary value into the input of the following gate.  That is for the $i^{th}$ component which is connected to the $i+1^{th}$ component, $output_i low \leq input_{i+1} low$ and $output_i high \geq input_{i+1} high$.  There is a naturally defined output threshold given the input thresholds.  That is passing the input threshold through the gate function G(). In addition, if the signal is monotonic then all inputs below the input low threshold or above the input high threshold gives valid output signals as well.  The reasoning is because for all $input low$ $x_i$ pairs, where $input low>x_i$ then $G(input low)>G(x_j)$ since G() is monotonic.  We also need to consider multiple input systems such as in biology where there can be multiple inputs to the same gate.  In biological systems where the ODE model is a sigmoidal curve in the form of a Hill function, the partial derivative with respect to any one input variable is always monotonic.  Thus this satisfies our monotonic requirement and allow for proper propagation.  In addition, because these equations are monotonic with regard to any one variable, there are only a limited number of logic that can be formed.  (expand?)

@@

The other property that sigmoidal curves are important for is the restorative property of signals to allow for noise and increase robustness.  

@@

@@

>>>>

\section{Correct boolean model}
In current designs of digital circuits, these are mostly synchronous circuits.  They are called synchronous due to the existence of a global clock which controls each logic gate.  The clock signals ensures that each gate are updated at the right time and also given enough time to propagate the correct output to the next gate. This avoids transmitting the wrong intermediate signal, which can cause an error in the final output compared to what we would expect.  There are also asynchronous circuit designs, where there is no such clocks and the variable delay times may cause one transition to happen before another at random. ()  Thus asynchronous systems must be designed more carefully to allow for these random events.  We expect biological systems to behave similarly to an asynchronous circuit because in biological systems there are no 'global clocks' that regulate the timing.  Because there is high variability between cells, and molecular interactions are of a stochastic nature, production from each gene element can have varying delay.  This variability differs from cell to cell and from one cycle to the next. Thus we would also expect a type of speed independence property to ensure that most of the time, the cells work reliably regardless of the random delays and advance through the cell cycle to produce new cells.
\newline \newline
From the bottom up, we extract properties from biological networks and use analogies with the well developed CMOS technology.  We start at the molecular level and demonstrate certain properties so that we can group gene elements together and represent them with a higher level boolean model.  We then show how this boolean model can be abstracted into a higher level forming building blocks that we can use in an asynchronous circuit. 
\newline \newline

The physics of gene protein interactions in the cell are congruent with properties that make it a valid digital circuit as well.  In past papers, gene regulatory networks has been modeled as boolean circuits with each gene/protein element represented by a boolean logic gate.  Each gene logic gate has inputs, such as activators and repressors, to determine whether the protein is being actively produced or not.  The mRNA production here can be modeled with the Hill function.  The sigmoidal shape of the Hill function is reminiscent of Voltage Transfer Characteristics of CMOS inverters.  As such, much of the design properties of CMOS circuits can be carried over.  There are two important properties that gates, which we use as building blocks in circuits, must satisfy.  The existence of a threshold, and the faithful propagation of a signal through a circuit, even when connected in a loop.  We examine these properties by first looking at the steady state solution of the equation for protein production.  The steady state solution is important because within a reasonable amount of time this is the output concentration of the protein given the inputs. This sigmoidal curve is then an input (the proteins that initiate transcription and production of the protein) versus the output, the current concentration of the protein.  The sigmoidal shape of the input vs output transfer curve during steady state allows for restoring property of the signal and hence faithful propagation in its next stages.   In an ODE model of the protein production and degradation with the Hill function, the steady state solutions are the Hill function divided by the degradation constant.  In general the protein production equation is of the form: $\frac{dprotein}{dt}=g(inputs)-protein\cdot\lambda$  where the function $g()$ is a sigmoidal curve such as the Hill function and $\lambda$ is the degradation constant.  It's steady state solution is $protein=g(inputs)/\lambda$, easily verifiable as also a sigmoidal Hill function.

@@

Thus we must verify the thresholds are upheld so that for all signals below the low threshold, it is read into the gate as a '0' and produces the correct corresponding output.  This must also be verified for all signals above the high threshold. We first work with a single input and make some assumptions:  1) that passing the signal at the low and high threshold though the gate produces the proper binary output signal, 2) the gate function is a monotonic function.  Then all inputs below the input low threshold or above the input high threshold gives valid output signals as well.  We first assume the function G() is monotonically increasing, the reasoning is thus for all $input low$ $x_i$ pairs, where $input low>x_i$ then $G(input low)>G(x_i)$.  The same could be shown for $input high$ and $x_j$.  If the function G() is monotonically decreasing then the inequality sign of the previous argument is flipped and the correct operation of the circuit still holds.  We verify that our initial assumptions hold in that we can set the output low and high thresholds accordingly, and the Hill function is indeed a monotonic function. 
%there are actually examples where this fails 0->0 in monotonically decreasing... need to fix too

@@

The next step is when gates are connected to each other, signals must propagate meaningfully as well.  The output of a gate should read as the same binary value into the input of the following gate based on their respective thresholds.  That is for the $i^{th}$ component which is connected to the $i+1^{th}$ component, $output_i low \leq input_{i+1} low$ and $output_i high \geq input_{i+1} high$.  Multiple inputs are common in biology where there can be activators and repressors and even degradation enzymes involved in one gene product.  This is easily reducible to the familiar problem in one dimension.  In biological systems where the ODE model is a sigmoidal curve in the form of a Hill function, the partial derivative with respect to any one input variable is always monotonic.  Thus this satisfies our monotonic requirement and allow for proper propagation. % In addition, because these equations are monotonic with regard to any one variable, there are only a limited number of logic that can be formed.  (expand?)
The other property that sigmoidal curves are important for is the restorative property of signals to allow for noise and increase robustness.  For the signals to allow a certain amount of noise to enter the system, we would want the output thresholds of one gate to be as far away from the input thresholds of the next gate as possible.  Thus when noise gets added to the output signal, it still has a good chance to transmit the correct binary data to the next stage.  In addition, these components can be connected in a loop.  In electronics, since the outputs and inputs are measured against the same unit of measurement, the voltage, the additional requirement is that the gain must be greater than 1.  This means that the slope of the function should be greater than 1.  Because we are using different chemical species at each stage to transmit the biological signal, we need to scale the slopes of the function accordingly.  

@@

Changing rates of degradation of proteins is an important aspect of the circuit that has been often ignored in previous models.  In past papers, gene regulatory networks has been modeled as boolean circuits with each gene/protein element represented by a boolean logic gate.  Each gene logic gate has inputs, usually its activators and repressors, to determine whether the protein is being actively produced or not.  These are usually modeled as AND or OR gates with its promoter binding compounds as its only inputs.  However, in our investigation using the ODE model of each gene, we found that degradation is a crucial part of the logic gates as well.  Protein degradation can be sped up through controlling the production of degradation enzymes.  Meanwhile for stable proteins, it is possible for the base degradation rate, the half life of the protein to be equivalent to the length of the cell cycle, the only reduction in protein concentration coming from the splitting of cells.  These proteins, once produced would remain in the cells for a very long time and essentially considered 'on' even when transcription and/or translation has stopped.  Thus there needs to be a separate way of representing the production of proteins from its degradation.  We take inspiration from Complementary Metal–Oxide–Semiconductor (CMOS) technology.  In CMOS there is a pull up network and a pull down network.  In the inverter example, the pull up network consists of a PMOS and the pull down network consists of a NMOS.  PMOS behaves, on first order approximation, as a wire when the input voltage is low and as an open circuit when the input voltage is high.  NMOS is the opposite, it behaves as an open when the input voltage is low and as a wire when the input voltage is high.  Thus when the input voltage is low, the output voltage is pulled up by the PMOS to a high voltage.  Meanwhile when the input voltage is high, the output voltage is pulled down to ground by the NMOS.  The transcriptional and translational regulation of these genes are similar to a pull up network while the accelerated degradation is similar to a pull down network.  If the background degradation is significantly slower, this effect can be thought of as the parasitic capacitance that slowly leak the charge of the output signal.  Overtime it degrades the signal however for short time scales it retains its voltage value and is a dynamic memory element.
(write about dynamic memory later?)
\includegraphics[width=\textwidth]{mosinaction}

@@

\section{Asynchronous circuits}
One important aspect of asynchronous circuit design is having memory elements.  This is because for a asynchronous circuit to be able to function, it must receive feedback signals from the next stage so that it knows if it needs to hold its current value, or if the next stage has received it's signal and can go to the next stage in processing.
